Learning how to act: making good decisions with machine learning

### 2017-08-03 thesis/thesis_map.tex
updated prior

@@

However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings. Getting them to generalise to even slightly different problems or data sets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

@@

\section{What makes a causal inference problem?}

@@

Identifying spam and automated essay marking both involve  processing text to determine an underlying (complex) attribute such as its topic or quality. In both cases, there is inherent competition between the algorithm and the people generating the text. As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label. Spammers and students will modify their writing in order to optimise their results. A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual. There are two key ways of ensuring this. The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm. The second is to ensure features are causes of the outcome and thus cannot be manipulated independtly of the label.

@@

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference. A predictive model may be sufficient if;

\begin{itemize}
\item Does acting on the predictions of the model change the mapping from features to target? (at least if the decision process is open to scrutiny). In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
\item 
This is more likely be be the case if the actions are recorded in the training data (unlike in the pneumonia example), and the algorithm has the same objective and information on which to base its decisions as generated the training data.

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes. 
\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? IE change people's behaviour.
\item Ethics - ... People's viewpoint on if its OK...
\end{itemize}

We can also view causal inference problems as a particular form of the transfer learning problem. We have training data obtained from one system, but we wish te estimate

It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.

Optimising hard to the wrong metric.

In conclusion, it is rare that anyone would build a model without indending that some kind of decion be made on the basis of its results. For this reason we assume that a causal model is required unless we can justify otherwise. I hope these examples gave you a feel for the richness and subtitles of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

@@

\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.

\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

\section{Approaches to causality}
This stuff gets built into explaining how the thesis fits together.

There are two broad approaches to deciding how to act. Reinforcement learning and causal inference. In reinforcement learning we estimate the effect of actions by taking them. We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward. This is a very powerful and general framework (REFERENCES TO GENERAL AI). However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING. We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.

We will call data sets where we do not have control over the decision making process that generated the data observational. Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control. There is a space in the middle where we have partially controlled the process by which agent select actions. Randomised data with imperfect compliance would be an example.

An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is

Reinforcement learning addresses the problem of learning from explicitly taking actions. There
is typically some state or environment. An agent chooses an action from those available in the
current state. The state then evolves stochastically as a function of the selected action and
the agent receives some feedback or reward that is a function of the new state. This setting
differs from the standard classification problem in that the agent must learn from feedback on
the selected action, rather than being presented with the correct action for a given state. A
common modelling assumption is that the state evolves only as a function of the previous state
and the action chosen, and given these, is independent of the previous history of states and
actions. This is known as a Markov decision process or MDP. A particularly well studied and
understood model is the singe state MDP. In this case, there are a set of actions, each associated
with a fixed but unknown reward distribution and at each time step our agent selects an action
and receives corresponding feedback. This is known as the multi-armed bandit problem.

Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
observational data. The key to causal inference is a framework that can model how actions
change the state of the world. This framework then allows us to map information collected in
one setting to another.

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
algorithms deal with the sequential nature of the decision making process, causal inference with
the problem that full randomisation is not always feasible, affordable or ethical. The similarities
between the problems that these techniques have been developed to address raises the question
of if there are problems best addressed by a combination of these approaches and how they
can be combined. The goal of my thesis is to explore these questions. In the next sections I
review the key literature in causal inference and bandits. I then present a general approach to
how causal models might be incorporated into bandit settings and conclude by demonstrating
an algorithm that leverages causal assumptions to improve performance in a specific bandit
setting.

There are two key approaches to causal problems. The first is to learn the outcome of actions by directly intervening in the system and seeing what happens. We then get feedback on how good those actions were. This is the approach taken in reinforcement learning. ADVANTAGES AND DISADVANTAGES OF THIS APPROACH. The second broad approach is causal inference. Here

\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}

\paragraph*{Two broad approaches} 
\begin{itemize}
\item Build a model to map the natural behaviour of the system to what will happen for some action
\item Take the action and see what happens
\end{itemize}

\paragraph*{The first is causal inference}

\paragraph*{The second is reinforcement learning}

\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions

Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments. Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation. Reinforcement learning studies how we can do better if the decisions are to be made sequentially.

\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.

\paragraph*{Limitations of causal inference}

\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?

@@

Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks. In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

@@

\begin{figure}[h]
\centering
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\end{tikzpicture}
\end{figure}

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$. Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution. Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

@@

\begin{figure}[h]
\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X \text{ (Treatment)}$};
\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
\path[]
	(2) edge (3);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}
\center
\caption{Causal Bayesian network for example \ref{exm:adjusting}}
\label{fig:causal_adjust}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]

 %nodes
\node[node](1){$Z \text{ (SES)}$};
\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3); 
\end{tikzpicture} 
\end{figure}

@@

\begin{figure}[h]
\center
\caption{Causal inference with ordinary Bayesian networks. The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$. The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$. As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate. However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
\label{fig:causal_ordinary_bayes}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
\tikzstyle{box}=[rectangle, draw=black!100]
 %nodes
\node[node](1){$Z$};
\node[node, below left=of 1](2){$X$};
\node[node, below right=of 1](3){$Y$};
\node[node, right=1.5cm of 3](4){$x$};
\node[lt, above right=of 4](5){$Z'$};
\node[lt, below right=of 5](6){$Y'$};
\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
\node[lt, above=0.5cm of 9](7){$\theta_z$};
\node[lt, above=2.0cm of 2](8){$\theta_x$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (6)
    (5) edge (6)
    (8) edge (2)
    (7) edge (1) edge (5)
    (9) edge (3) edge (6);
\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
\end{tikzpicture} 
\end{figure}

@@

\begin{figure}[h]
\caption{Randomised treatment with imperfect compliance}
\label{fig:randomized_imperfect_compliance}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Prescribed Treatment};
\node[node, below right=of 1](2){Actual Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above left=of 3](4){U};
\path[]
	(1) edge (2)
	(2) edge (3) 
	(4) edge (3) edge (2);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\captionsetup[subfigure]{position=b}
\centering
\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
\label{fig:metaphysical_distribution_difference}
\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$. The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated. \label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
\hspace{0.02\textwidth}
\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
\end{figure}

@@

\begin{figure}[h]
\caption{SEM for example \ref{exm:ranomized_experiment}}
\label{fig:sem_randomized_treatment}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(\epsilon_x)$};
\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\caption{SEM for example \ref{exm:adjusting}}
\label{fig:sem:preschool}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(Z,\epsilon_x)$};
\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
\node[lt, above=of 6](7){$\epsilon_z$};
\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2)
	(7) edge (6)
	(6) edge (2) edge (3);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
\label{fig:simpsons_paradox}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Gender};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}[h]
\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
\label{fig:simpsons_paradox2}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Blood Pressure};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\end{figure}

@@

\begin{figure}
\centering
\caption{Simpson's reversal visualised. The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors. However, when we sum them, the ratio is steeper for the un-primed variables.}
\label{fig:simpsons_vector}
\includegraphics[scale=.4]{figures/simpson_vector}
\end{figure}

@@

\end{tikzpicture}
\end{subfigure}
\end{figure}

@@

>>>>

\section*{Questions}
\begin{itemize}
\item Are there any problematic counterfactuals being used in defining bandit problems
\item in the Schölkopf causal fairness paper - is there an issue with the use of twin networks. Could it be fixed with SWIGs?
\item Contextual bandits are solving this argmax h problem. Should causal inference actually be solving the same thing?
\item what is the connection between which variables should be included when doing estimating causal effects from observational data and stratification in randomized experimental design?
\item I don't understand why covariate shift is not supposed to be an issue in Bayesian models. 
\item What does conditioning mean in the Bayesian sense? - can imposing a prior which is true with respect to that variable, increase bias due to confounding? (In the same way that in the pnemonia case, the coefficient for Asthma should be set to zero after the model has been fit, not before.

\item Does the way we define supervised learning and loss fuctions imply an additive noise model? Why would we expect this to hold - why don't we run into more problems?

\end{itemize}

@@

However, despite the huge advances in machine learning (in particular deep learning), machine learning algorithms are effective only within narrow problem settings. Getting them to generalise to even slightly different problems or data sets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

@@

\section{What makes a problem causal?}

@@

Identifying spam and automated essay marking both involve  processing text to determine an underlying (complex) attribute such as its topic or quality. In both cases, there is inherent competition between the algorithm and the people generating the text. As a result, decisions made by the algorithm are likely to change the relationship between the features it relies on and the true label. Spammers and students will modify their writing in order to optimise their results. A standard supervised learning approach can only work if the resulting change in the mapping from features to label is sufficiently gradual. There are two key ways of ensuring this. The first is to limit people's ability to observe (and thus react to) decisions made by the algorithm. The second is to use a model in which the features are related to the outcome in such a way that they cannot be manipulated independently.

@@

Having considered these examples we can now identify some general aspects of problems that require causal (as opposed to purely predicive) inference. A predictive model may be sufficient if, given the variable(s) being predicted, it is clear which action is optimal and if selecting actions on the basis of the model does not change the mapping from features to outcomes. The second requirement is particularly difficult to satisfy when an algorithm is making important decisions effecting individual people. Think about problems like credit scoring and parole decisons. There are strong ethical grounds for demanding transparancy, but if the goals of society and the individuals are not perfectly aligned and there is any posibility that people can manipulate features independtly of the outcome, there will be a conflict between model accuracy and transparancy. It is rare to build a model without any intent to make some kind of decion on the basis of its results. Thus, I argue we should assume a causal model is required until we can justify otherwise.

@@

Causal questions are everywhere. Statistics, economics, social science, artificial intelligence and machine learning. Techniques and for addressing such problems have developed in parralel within many diciplines. These techniques can usefully be categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there are a set of available actions, the agent repeately decides which to select and then observes the outcome of the chosen action. They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.

The approach of learning the outcome of an action by taking it has plays a key role in advancing our knoweledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the affect of an action from such observational data sets is the problem addressed by observational causal inference.

Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the orignial system. This is impossible without some assumptions about how the the orignal system and the system after intervention are related to one-another. The key to observational inference is a framework that models how actions change the state of the world and thus allows us to map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem randomisation is not always feasible, affordable or ethical. The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined.

The goals of this thesis are: to clarify the distinction between prediction and causal inference and highlight when and how we can leverage advances in prediction to improve causal inference and to draw together the key approaches to solving causal problems, from both the interventionalist and observational viewpoints. In chapter two, I describe the existing frameworks for formalising causality developed withing economics, statistics and machine learning and discuss how they relate to one another. Chapter three reviews the key approaches to causal inference from observational data. Chapter four highlights the interventionalist viewpoint, covering traditional randomised experiments and the closely related multi-armed bandit problems and highlights connections between the algorithms applied to bandit problems and those for estimating causal effects from observational data.

In chapter five, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

Chapters two to four predominantly review the existing liturature with the goal of making connections between the different viewpoints and approaches. They do not contain novel technical results but present the material in a unified way. Chapter five is a novel contribution. It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit liturature and describes an algorithm that can leverage this structure to make better decisons.

@@

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks. In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]

@@

\end{tikzpicture}
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={observed}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\caption{Some valid Bayesian networks for a distribution $P$ over $(X_1,X_2,X_3)$ in which $X_3$ is conditionally independent of $X_1$ given $X_2$, denoted $X_3 \ci X_1 | X_2$. Graphs (a), (b) and (c) are all a \emph{perfect map} for $P$ as the graphical stucture implies exactly the same set of independencies exhibitied by the distribution. Graph (d), like figure \ref{fig:bayesnet} does not imply any conditional independencies, and is thus a valid (but not very useful) Bayesian network representation for any distribution over three variables.}
\label{fig:bayesnet2}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X \text{ (Treatment)}$};
\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
\path[]
	(2) edge (3);
\end{tikzpicture}
\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example}
\end{figure}

@@

\begin{figure}
\center
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]

 %nodes
\node[node](1){$Z \text{ (SES)}$};
\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3); 
\end{tikzpicture}
\caption{Causal Bayesian network for example \ref{exm:adjusting}}
\label{fig:causal_adjust} 
\end{figure}

@@

\begin{figure}[h]
\center
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observed},lt/.style={latent}]
\tikzstyle{box}=[rectangle, draw=black!100]
 %nodes
\node[node](1){$Z$};
\node[node, below left=of 1](2){$X$};
\node[node, below right=of 1](3){$Y$};
\node[node, right=1.5cm of 3](4){$x$};
\node[lt, above right=of 4](5){$Z'$};
\node[lt, below right=of 5](6){$Y'$};
\node[lt, above=2.0cm of 3](9){$\theta_{y}$};
\node[lt, above=0.5cm of 9](7){$\theta_z$};
\node[lt, above=2.0cm of 2](8){$\theta_x$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (6)
    (5) edge (6)
    (8) edge (2)
    (7) edge (1) edge (5)
    (9) edge (3) edge (6);
\node[rectangle, inner sep=4.0mm, fit= (1) (2) (3),draw=black!100,label=below right:N] {};
\node[rectangle, inner sep=4.0mm, fit= (5) (6),draw=black!100,label=below right:M] {};
\end{tikzpicture} 
\caption{Causal inference with ordinary Bayesian networks. The plate on the left represents the observed data generated prior to the intervention and the plate on the right the data we anticipate obtaining after an intervention that sets the pre-interventional variable $X$ to $x$. The assumptions characterised by this plate model correspond to those implied by the causal Bayesian network in figure \ref{fig:causal_adjust} for the intervention $do(X=x)$. As the networks in this figure are ordinary Bayesian networks, we could have represented the same information with a different ordering of the links within each plate. However, we would then have a complex transformation relating the parameters between the two plates rather than a simple invariance.}
\label{fig:causal_ordinary_bayes}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Prescribed Treatment};
\node[node, below right=of 1](2){Actual Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above left=of 3](4){U};
\path[]
	(1) edge (2)
	(2) edge (3) 
	(4) edge (3) edge (2);
\end{tikzpicture}
\caption{Randomised treatment with imperfect compliance}
\label{fig:randomized_imperfect_compliance}
\end{figure}

@@

\begin{figure}[h]
\captionsetup[subfigure]{position=b}
\centering
\subcaptionbox{Marginal distributions over the potential outcomes $\cf{Y}{1}$ and $\cf{Y}{0}$ for $\mu_1=1$, $\mu_0=0$ and $\sigma = 1$. The blue curve shows the distribution of $Y$ if everyone were to be treated and the blue curve the distribution if no-one was treated. \label{fig:metaphys_a}}{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_a.pdf}}
\hspace{0.02\textwidth}
\subcaptionbox{Two very different distributions of individual causal effects consistent with the potential outcome distributions. \label{fig:metaphys_b}}
{\includegraphics[width=.45\linewidth]{figures/counterfactual_nonidentify_b.pdf}}
\caption{The distribution of individual treatment effects is not identifiable, even from a randomised controlled trial.}
\label{fig:metaphysical_distribution_difference}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(\epsilon_x)$};
\node[node, right=of 2](3){$Y = f_y(X,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2);
\end{tikzpicture}
\caption{SEM for example \ref{exm:ranomized_experiment}}
\label{fig:sem_randomized_treatment}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X = f_x(Z,\epsilon_x)$};
\node[node,above right=of 2](6){$Z = f_z(\epsilon_z)$};
\node[lt, above=of 6](7){$\epsilon_z$};
\node[node, below right=of 6](3){$Y = f_y(X,Z,\epsilon_y)$};
\node[lt, above=of 3](4){$\epsilon_y$};
\node[lt, above=of 2](5){$\epsilon_x$};
\path[]
	(2) edge (3)
	(4) edge (3)
	(5) edge (2)
	(7) edge (6)
	(6) edge (2) edge (3);
\end{tikzpicture}
\caption{SEM for example \ref{exm:adjusting}}
\label{fig:sem:preschool}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Gender};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (2) edge (3) 
	(2) edge (3);
\end{tikzpicture}
\caption{An example of causal network that can give rise to Simpson's Paradox. In this case we should select treatment on the basis of the gender specific results.}
\label{fig:simpsons_paradox}
\end{figure}

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](1){Z = Blood Pressure};
\node[node, below left=of 1](2){T = Treatment};
\node[node, below right=of 1](3){Y = Outcome};
\path[]
	(1) edge (3) 
	(2) edge (3)
	(2) edge (1);
\end{tikzpicture}
\caption{Another causal network that can exhibit Simpson's paradox. In this case, \quotes{the solution} is not to stratify on $Z$.}
\label{fig:simpsons_paradox2}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=.4]{figures/simpson_vector}
\caption{Simpson's reversal visualised. The ratios involving $N_i'$ are steeper than than those involving $N_i$ for both the blue and green vectors. However, when we sum them, the ratio is steeper for the un-primed variables.}
\label{fig:simpsons_vector}
\end{figure}

@@

\end{tikzpicture}
\end{subfigure}
\caption{Rule 1 example. $(Y \ci W|X)$ in  $G_{\overline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|do(X)}$ }
\label{fig:mutilatednet}
\end{figure}

@@

\chapter*{Careful with that prior}
